Constraining the Interiors of Asteroids Through Close Encounters

### 2022-04-21 paper/main.tex
Reviewed paper

% Only include extra packages if you really need them. Common packages are:
\usepackage{graphicx}	% Including figure files
\usepackage{amsmath}	% Advanced maths commands
\usepackage{amssymb}	% Extra maths symbols
\usepackage{xcolor}
\usepackage{tikz}
\usepackage{colortbl}
\usepackage{dsfont}
\usepackage{bm}

@@

% Abstract of the paper
\begin{abstract}
  Knowledge of the interior distribution of an asteroid can reveal its composition and constrain its evolutionary history. However, most observations are not sensitive to interior structure. We investigate the degree to which asteroid interior density distributions can be extracted from rotational velocity data gathered during a close encounter. We derive the equations of motion for an rigid asteroid's orientation and angular velocity to arbitrary order and use them to generate synthetic rotational velocity data for a representative asteroid on a close Earth encounter. Using Markov Chain Monte Carlo fits, we re-extract the density moments of the asteroid in a wide range of scenarios to measure the degree to which best-fitting parameter precision is affected. Specifically, we use many injection-retrieval tests to study fit precision's dependence on the asteroid's moment of inertia, observational precision and cadence, orbital parameters, and initial spin pole direction as well as the quantity of near-pericentre data and the central body oblateness. Finally, we discuss the degeneracy between the density moments and the actual density distribution and propose three models to construct a representative density distribution from fit results.
\end{abstract}

@@

Over the past twenty years, the increase in quantity and quality of sensitive all-sky surveys has prompted the discovery of numerous asteroids. Such advances have been made via ground-based surveys such as the Catalina Sky Survey \cite{larson1998catalina}, Pan-STARRS \cite{kaiser2002pan}, and the Lincoln Near-Earth Asteroid Research project (LINEAR) \cite{stokes2000lincoln}, as well as space-based instruments such as the Wide-field Infrared Survey Explorer (WISE) mission \cite{wright2010wide}. Many of these asteroids are relatively small, but some are kilometre-sized and a few are predicted to encounter Earth or other planets closely in the near future. More encounter candidates are likely to be discovered in the future --- for Earth or other planets --- potentially by the Large-aperture Synoptic Survey Telescope (LSST) or global telescope networks such as the Las Cumbres Observatory (LCO). Light curve analysis has been performed on many of these new asteroids to study their surface properties, but observations that are sensitive to the interior structure of the asteroid are rarer.

However, tidal torque acting on an asteroid and perturbing its angular velocity during an encounter presents an opportunity to constrain the asteroid mass distribution. Such angular velocity perturbations have been observed and used to extract asteroid properties in several cases, including for the 2013 encounter of (367943) Duende with Earth \cite{MOSKOVITZ2020113519, benson2020spin}, and asteroid binaries (3905) Doppler and (617) Patroclus respectively. Altered rotational states altered by Solar tidal torque have been excluded for 1I/2017 U1 `Oumuamua \cite{KWIECINSKI2018170}.

The upcoming encounter of Apophis is of particular interest and much theoretical work has been devoted to accurate estimates of its orbital and physical properties, and predictions of how its angular velocity will be affected by tidal torque \cite{yu2014numerical, hirabayashi2021finite,valvano2022apophis, Lee2022Apophis, souchay2014rotational, souchay2018changes}. This research builds on a larger body of theoretical work showing that the tidal torque, observed through angular velocity perturbations, is sensitive to asteroid interior density distribution \cite{Naidu_2015, Makarov2022ChaosOO, RICHARDSON199847, scheeres2004evolution}. This two-body generalized system has been studied to different orders and with several different methods \cite{paul88, SCHEERES2000106, ashenberg07, BOUE2009750, HouMar2017}. The success of this work is demonstrated by the recent extraction of moment of inertia ratios for Duende \cite{MOSKOVITZ2020113519}. However, density distribution information beyond the moment of inertia ratios have not yet been extracted for planetary encounters. More research is needed to study in what cases these effects are observable, and what factors generally inhibit observation of these sub-dominant terms.

In this paper, we address this question by deriving a new equation of motion for an asteroid encounter system, using , and describe a simulation used to integrate the equation of motion and produce synthetic data of angular velocity over time (section \ref{sec:sim}).it to simulate a generic asteroid encounter of Earth. In section  \ref{sec:methods}, we then extract physical characteristics of the asteroid (specifically, density moments) from synthetic data via a Markov Chain Monte Carlo (MCMC) fit. We also describe methods to extract estimates of density distribution from these density moments. In section \ref{sec:results}, we demonstrate our fit process and evaluate our density distribution extraction methods on density moments extracted from fits to synthetic data. In section \ref{sec:discussion}, we measure the uncertainties in the Posterior Probability Distribution (PPD) of density moments as a function of various encounter parameters such as the observational uncertainty, encounter orbital parameters, asteroid physical properties, and more.

@@

In appendix \ref{app:eom}, we derive a novel equation for the tidal torque experienced by an asteroid in a two-body system, accurate to arbitrary order. The equation is written in terms of the asteroid ``length'' 
\begin{equation}
  a_\mathcal{A}^2 = \frac{1}{\mu_\mathcal{A}} \int_\mathcal{A} d^3 r \rho_\mathcal{A}(\bm r) r^2
  \label{eqn:am}
\end{equation}
and (complex) ``density moments''
\begin{equation}
  K_{\ell m} = \frac{1}{\mu_\mathcal{A} a_\mathcal{A}^\ell} \int_\mathcal{A} d^3 r \rho_\mathcal{A}(\bm r) R_{\ell m}(\bm r),
  \label{eqn:klm}
\end{equation}
both of which are constant and only need to be computed once to generate data. Here, $\rho_\mathcal{A}(\bm r)$ is the asteroid density distribution, $\mu_\mathcal{A}$ is the asteroid mass, and $R_{\ell m}$ are the regular solid spherical harmonics (see the appendix for details). They produce a tidal torque of 
\begin{equation}
  \begin{split}
  \bm \tau = & G\frac{\mu_\mathcal{A}\mu_\mathcal{B}}{2}\left[\sum_{\ell m} a_\mathcal{B}^\ell J_{\ell m} \sum_{\ell' m'}a_\mathcal{A}^{\ell'}S^*_{\ell+\ell', m + m'} (\bm D) (-1)^{\ell'}\right.\\
  \times & \left.\sum_{m''=-\ell'}^{\ell'} \sqrt{\frac{(\ell'-m'')!(\ell'+m'')!}{(\ell'-m')!(\ell'+m')!}}  \mathcal{D}^{\ell'}_{m'm''}(\alpha, \beta, \gamma)^* \right. \\
  \times & \Big((i\unit x - \unit y)(\ell'-m''+1)K_{\ell',m''-1} \\
  & +(i\unit x+\unit y)(\ell'+m''+1)K_{\ell',m''+1}+2im''\unit z K_{\ell'm''}\Big) \Bigg],
  \end{split}
  \label{eqn:tidal-torque}
\end{equation}
where $\bm D$ is the position of the asteroid; $\alpha$, $\beta$, and $\gamma$ are Euler angles expressing the orientation of the asteroid; and  $\mu_\mathcal{B}$ and $J_{\ell m}$ are the mass and density moments of the central body respectively; and $S_{\ell m}$ are the irregular solid spherical harmonics.

In section \ref{sec:sim}, we describe the simulation used to integrate the equations of motion. In section \ref{sec:uncertainty}, we define the uncertainty model used to add noise to the data, and we describe how PPDs for density moments were extracted from spin-pole data in section \ref{sec:fit}. Finally, we outline a general model to extract density distributions from density moments in section \ref{sec:general-density} and define specific model implementations in section \ref{sec:specific-density}

@@

We built a custom simulation in C++ to produce angular velocity data as a function of time. This simulation requires as initial data (1) the orbital parameters of the asteroid $r_p$ (perigee distance) and $v_\infty$ (hyperbolic excess velocity); (2) the cadence of angular velocity observation $\Delta t$; (3) the central body moments $J_{\ell m}$, mass $\mu_\mathcal{B}$, and radius $a_\mathcal{B}$; (4) the initial asteroid angular velocity in the inertial frame $\bm \Omega_0$; (5) the asteroid length $a_\mathcal{A}$; and (6) the asteroid's density moments $K_{\ell m}$ and initial Euler angle $\gamma_0$. All parameters except (6) are assumed to be known to high accuracy. One can imagine that $a_\mathcal{A}$ is approximated by light-curve analysis, but if not, it is still necessary to fix $a_\mathcal{A}$ or else the values of $K_{\ell m}$ are degenerate with $a_\mathcal{A}$.

@@

With the simulation inputs specified, the equations of motion are integrated via the Runge-Kutta fourth order method, with a variable time step
\begin{equation}
  dt = dt_\text{min} + 10^{-3}(dt_\text{max} - dt_\text{min}) \brackets{\parens{\frac{D}{r_p}}^3 - 1}.
\end{equation}
The parameters $dt_\text{max}$ and $dt_\text{min}$ (20 and 10 seconds respectively) were chosen such that the error was $\sim$100 times the floating point error, and that neighbouring values of $K_{\ell m}$ yielded significantly different spin pole data compared to floating point error. The data used to choose this $dt$ was obtained using the reference asteroid configuration, described in appendix \ref{app:reference-configs}.

@@

To add noise to data generated via the above simulation, we use the following uncertainty model. Each asteroid spin vector $\bm \Omega$ is assumed to be uncorrelated with other spin vectors, and we model uncertainty in the orientation and in the period as also uncorrelated. Consider a true spin vector $\bm \Omega^*$. For the sake of description, we work in coordinates in which $\bm \Omega^* \parallel \unit z$. Then, expressing the observed spin vector $\bm \Omega$ in spherical coordinates, we draw the polar angle from a normal distribution with standard deviation $\sigma_\theta$ centred on zero and the azimuthal angle from a uniform distribution. We also draw the ratio $\rho=\Omega/\Omega^*$ from a log-normal distribution centred on one, with width $\sigma_\rho$. We generally represent $\sigma_\theta > \sigma_\rho$ under the assumption that period is better constrained by light-curve data than spin pole. Explicitly, the probability density function (PDF) of $\rho$ is 
\begin{equation}
  P(\rho) = \frac{1}{\rho\sqrt{2\pi \sigma_\rho^2}} \exp\parens{-\frac{\ln^2\rho}{2\sigma_\rho^2}}.
\end{equation}
See figure \ref{fig:uncertainty-model} for an illustration of the uncertainty model.

@@

  \draw[-{Latex[length=3mm]}] (0, 0) -- (-1.25, 3.35) node[anchor=south west] {$\bm \Omega$};
  \draw[rotate around={-50:(-3,3)}] (-3,3) ellipse (1.4 and 2);
  \draw (-3, 3) arc (110:93:6);
  \draw (-2, 3.5) node[anchor=center] {$\theta$};
  \draw (-1, 1.8) node[anchor=center] {$\rho \Omega^*$};

@@

\subsection{Extracting density moments from spin data}
\label{sec:fit}
Given synthetic data, an Affine Invariant Markov Chain Monte Carlo (MCMC) Ensemble sampler was used to generate PPDs from flat priors. We use the Python implementation \texttt{emcee} \cite{foreman2013emcee}. Our parameters were $\gamma_0$, $K_{20}$, $K_{22}$, and $K_{3m}$ (10 in total), and were bounded by $|\gamma_0| < \pi/4$, and bounds on $K_{2 m}$ given in equation \ref{eqn:parameter-bounds}. Note that $\gamma_0$ is degenerate with $\gamma_0 + \pi/2$ since this is equivalent to re-labeling $\unit y$ as $\unit x$ and $\unit x$ as $-\unit y$. The other bounds were $|K_{3m}| < 1$.

The MCMC was determined to converge when the fractional change in autocorrelation time (computed every 100 iterations) was one percent, and the number of iterations computed so far was more than 100 times the autocorrelation time. The MCMC fit also was set to terminate if more than $10^5$ iterations were run, but this only occurred for fits in which the data was too low quality to resolve the parameters. $~10^4$ iterations was often sufficient, which generally consumed about 7 hours of computation time on a super computer running 16 threads on 8 cores.

Before the MCMC was run, local minima in the likelihood were found via the Nelder-Mead algorithm implemented in \texttt{scipy} \cite{Gao2012}. It was found that only one local minimum existed, except when $K_{22}=0$ in which case rotational symmetry caused multiple values of $\gamma_0$ to be degenerate. Walkers were initialized near this local minimum, distributed with probability distributed as a Gaussian approximation of the likelihood, as determined via the inverse Hessian of the likelihood at the minimum. Due to the high sensitivity of the angular velocity data to density moments, the minimization procedure sometimes failed to isolate the minimum likelihood. Therefore, a simpler simulation without the $K_{3m}$ terms of equation \ref{eqn:tidal-torque} was first used to minimize likelihood as a function of $\gamma_0$ and $K_{2m}$, and then the full simulation was used to find $K_{3m}$, with $\gamma_0$ and $K_{2m}$ fixed. This tiered minimization process motivates us to call $\gamma_0$ and $K_{2m}$ the ``first-order parameters'' and $K_{3m}$ the ``second-order parameters.''

We further subdivided the minimization process by first minimizing with respect to data truncated at the point after perigee where $D=fr_p$ for some manually-set fraction $f$. In practice, $f=2$ often led to success. The minimum was then further refined by minimizing based on the full data, with the previous minimum as the initial estimate.

@@

The asteroid density distribution $\rho_\mathcal{A}(\bm r)$ is not uniquely constrained via tidal torque interactions because only the density moments $K_{\ell m}$ contribute to equation \ref{eqn:tidal-torque}. However, by making sufficient assumptions about the density distribution, we can nevertheless estimate $\rho(\bm r)$ from $K_{\ell m}$. Here, we outline three different assumptions which constitute three models for extracting density distributions from density moments. One of them can be chosen, or another created, depending on what features are desired to be represented in the density distribution.

@@

We define a new coordinate system, the ``hybrid frame,'' which coincides exactly with body-fixed frame at the initial orientation of the asteroid assuming that the fit result for $\gamma_0$ is perfectly accurate. The orientation of the hybrid frame with respect to the inertial frame is therefore exactly known, so that $\mathds{1}$ is also exact in the hybrid frame. The $K_{\ell m}$ components are known in the body-fixed frame which aligns with the hybrid frame up to uncertainty in $\gamma_0$. We will solve for $\rho(\bm r)$ in the hybrid frame given fit results $K_{\ell m}$ (known in the body-fixed frame) and the fixed $a_\mathcal{A}$ (equal in all frames).

The density moments defined in equation \ref{eqn:klm} are not linear in $\rho(\bm r)$, but we may fix $a_\mathcal{A}$ at the value assumed during the fit, removing the non-linearity induced by division by $a_\mathcal{A}$. Furthermore, since the system is independent of the total asteroid mass, we set $\mu_\mathcal{A}$ equal to an arbitrary constant which makes equation \ref{eqn:klm} linear in $\rho(\bm r)$. Similarly, the form of equation \ref{eqn:am} guarantees that $a_\mathcal{A}^2$ is linear in $\rho(\bm r)$. Equation \ref{eqn:am} and the $K_{00}=1$ component of equation \ref{eqn:klm} can each be applied as constraints to enforce these choices of $a_\mathcal{A}^2$ and $\mu_\mathcal{A}$.

Suppose that we restrict the number of degrees of freedom of $\rho(\bm r)$ from infinity to $m$ by explicitly defining some function $\rho(\bm r, \bm \Theta)$ for an $m$-dimensional vector $\bm \Theta$ which contains the free parameters of $\rho$. We will leave the explicit definition of $\rho(\bm r, \bm \Theta)$ to the model descriptions below, but for now we assume that $\rho$ is linear in $\bm \Theta$; i.e.,
\begin{equation}
  \rho(\bm r, \bm \Theta) = \bm B(\bm r) \cdot \bm \Theta
  \label{eqn:density-distro}
\end{equation}
for a $m$-dimensional vector $\bm B(\bm r)$. ($\bm B$ need not be linear in $\bm r$.) Thus, the defining equations of $a_\mathcal{A}^2$ and $K_{\ell m}$ are linear in $\bm \Theta$. We further assume that the model describes a way to reverse this equation, to write
\begin{equation}
  \bm \Theta = A \bm K
  \label{eqn:density-model}
\end{equation}
where $A$ is a matrix. Here, we have arranged $a_\mathcal{A}^2$ and $K_{\ell m}$ into a vector $K$, which we say has $n$ dimensions. The order of this arrangement is irrelevant, as long as it is kept consistent.

To propagate uncertainties from $\bm K$ to $\bm \Theta$ and therefore $\rho(\bm r)$, we need the covariance matrix $\Sigma_K$ for $\bm K$. First, assume that the hybrid frame is offset from the body-fixed frame by some small angle $\Delta \gamma$, which results from uncertainty in $\gamma_0$. Then
\begin{equation}
  K_{\ell m}^\mathrm{hybrid} = e^{-im\Delta \gamma}K_{\ell m}^\mathrm{body-fixed}.
  \label{eqn:body-fixed-to-hybrid}
\end{equation}
by equation \ref{eqn:ylm-rotation}. Since $K_{\ell m}^\mathrm{body-fixed}$ was obtained by an MCMC fit, a large set of PPD-distributed samples is available for $\Delta \gamma$ and $K_{\ell m}^\mathrm{body-fixed}$, and the covariance matrix $\Sigma_K$ can be computed statistically in the hybrid frame by applying equation \ref{eqn:body-fixed-to-hybrid} to the samples. Then, propagation of uncertainty guarantees that the covariance matrix of $\bm \Theta$ is $\Sigma_\Theta = A \Sigma_K A^*$ and the density distribution and uncertainty on density distribution are equal to
\begin{equation}
  \rho(\bm r) = \bm B(\bm r)^* A\bm K \qquad \sigma^2_\rho(\bm r) = \bm B(\bm r)^* A \Sigma_K A^* \bm B(\bm r).
  \label{eqn:unc-rho}
\end{equation}
where $A^*$ is the adjoint of $A$.

>>>>

% Only include extra packages if you really need them. Common packages are:
\usepackage{graphicx}	% Including figure files
\usepackage{amsmath}	% Advanced maths commands
\usepackage{amssymb}	% Extra maths symbols
\usepackage{xcolor}
\usepackage{tikz}
\usepackage{colortbl}
\usepackage{dsfont}
\usepackage{siunitx}
\usepackage{bm}

@@

% Abstract of the paper
\begin{abstract}
  Knowledge of the interior distribution of an asteroid can reveal its composition and constrain its evolutionary history. However, most asteroid observational techniques are not sensitive to interior structure. We investigate the degree to which interior density distributions can be extracted from rotational velocity data gathered during a close encounter. We derive the equations of motion for a rigid asteroid's orientation and angular velocity to arbitrary order and use them to generate synthetic rotational velocity data for a representative asteroid on a close Earth encounter. Using Markov Chain Monte Carlo fits, we re-extract the density moments of the asteroid in a wide range of scenarios to measure the degree to which best-fitting parameter precision is affected. Specifically, we use many injection-retrieval tests to study fit precision's dependence on the asteroid's moment of inertia, observational precision and cadence, orbital parameters, and initial spin pole direction as well as the planet being encountered, the quantity of near-pericentre data, and the central body oblateness. Finally, we discuss the degeneracy between the density moments and the actual density distribution and propose three models to construct a representative density distribution from fit results.
\end{abstract}

@@

Over the past twenty years, the increase in quantity and quality of sensitive all-sky surveys has prompted the discovery of numerous asteroids. Such advances have been made via ground-based surveys such as the Catalina Sky Survey \cite{larson1998catalina}, Pan-STARRS \cite{kaiser2002pan}, and the Lincoln Near-Earth Asteroid Research project (LINEAR) \cite{stokes2000lincoln}, as well as space-based instruments such as the Wide-field Infrared Survey Explorer (WISE) mission \cite{wright2010wide}. Many of these asteroids are relatively small, but some are kilometre-sized and a few are predicted to encounter Earth or other planets closely in the near future. More encounter candidates are likely to be discovered in the future by new efforts such as the Large-aperture Synoptic Survey Telescope (LSST) or the Las Cumbres Observatory (LCO). Light curve analysis has been performed on many of these new asteroids to study their surface properties, but observations that are sensitive to the interior structure of the asteroid are rarer.

However, tidal torque acting on an asteroid and perturbing its angular velocity during an encounter presents an opportunity to constrain the interior mass distribution. Such angular velocity perturbations have been observed and used to extract asteroid properties in several cases, including for the 2013 encounter of (367943) Duende with Earth \cite{MOSKOVITZ2020113519, benson2020spin}, and asteroid binaries (3905) Doppler and (617) Patroclus respectively. Altered rotational states altered by Solar tidal torque have been excluded for 1I/2017 U1 `Oumuamua \cite{KWIECINSKI2018170}.

The upcoming encounter of Apophis is of particular interest and much theoretical work has been devoted to accurate estimates of its orbital and physical properties, and predictions of how its angular velocity will be affected by tidal torque \cite{yu2014numerical, hirabayashi2021finite,valvano2022apophis, Lee2022Apophis, souchay2014rotational, souchay2018changes}. This research builds on a larger body of theoretical work showing that the tidal torque, observed through angular velocity perturbations, is sensitive to asteroid interior density distribution \cite{Naidu_2015, Makarov2022ChaosOO, RICHARDSON199847, scheeres2004evolution}. This two-body generalized system has been studied to different orders and with several different methods \cite{paul88, SCHEERES2000106, ashenberg07, BOUE2009750, HouMar2017}. One success of this work was the recent extraction of moment of inertia ratios for Duende \cite{MOSKOVITZ2020113519}. However, density distribution features beyond the moment of inertia ratios have not yet been extracted for any asteroid encounters. More research is needed to study in what cases these effects are observable, and what factors generally inhibit observation of these new features.

In this paper, we address this question by deriving a new equation of motion for an asteroid encounter system which is quickly computable and valid to arbitrary order. In section \ref{sec:methods}, we describe a simulation used to integrate the equations of motion and produce synthetic data of angular velocity over time, followed by a Markov Chain Monte Carlo (MCMC) fit process which  extracts physical characteristics of the asteroid (specifically, density moments) from the fit data. We then describe three methods to generate full density distributions from the density moments. Section \ref{sec:results} demonstrates the efficacy of this spin-data-to-density-distribution pipeline and explores the key properties. Finally, in section \ref{sec:discussion}, we measure the uncertainties of the best-fitting parameter estimates as a function of various encounter properties to determine which properties most inhibit accurate extraction of density distributions.

@@

Only some aspects of an asteroid's density distribution affect tidal torque interactions. These are the asteroid ``length'' 
\begin{equation}
  a_\mathcal{A}^2 = \frac{1}{\mu_\mathcal{A}} \int_\mathcal{A} d^3 r \rho_\mathcal{A}(\bm r) r^2
  \label{eqn:am}
\end{equation}
and (complex) ``density moments''
\begin{equation}
  K_{\ell m} = \frac{1}{\mu_\mathcal{A} a_\mathcal{A}^\ell} \int_\mathcal{A} d^3 r \rho_\mathcal{A}(\bm r) R_{\ell m}(\bm r),
  \label{eqn:klm}
\end{equation}
both of which are constant and only need to be computed once to generate data. Here, $\rho_\mathcal{A}(\bm r)$ is the asteroid density distribution, $\mu_\mathcal{A}$ is the asteroid mass, and $R_{\ell m}$ are the regular solid spherical harmonics (see the appendix for details). They produce a tidal torque of 
\begin{equation}
  \begin{split}
  \bm \tau = & G\frac{\mu_\mathcal{A}\mu_\mathcal{B}}{2}\left[\sum_{\ell m} a_\mathcal{B}^\ell J_{\ell m} \sum_{\ell' m'}a_\mathcal{A}^{\ell'}S^*_{\ell+\ell', m + m'} (\bm D) (-1)^{\ell'}\right.\\
  \times & \left.\sum_{m''=-\ell'}^{\ell'} \sqrt{\frac{(\ell'-m'')!(\ell'+m'')!}{(\ell'-m')!(\ell'+m')!}}  \mathcal{D}^{\ell'}_{m'm''}(\alpha, \beta, \gamma)^* \right. \\
  \times & \Big((i\unit x - \unit y)(\ell'-m''+1)K_{\ell',m''-1} \\
  & +(i\unit x+\unit y)(\ell'+m''+1)K_{\ell',m''+1}+2im''\unit z K_{\ell'm''}\Big) \Bigg],
  \end{split}
  \label{eqn:tidal-torque}
\end{equation}
where $\bm D$ is the position of the asteroid; $\alpha$, $\beta$, and $\gamma$ are $z-y-z$ Euler angles expressing the orientation of the asteroid; $\mu_\mathcal{B}$ and $J_{\ell m}$ are the mass and density moments of the central body respectively; and $S_{\ell m}$ are the irregular solid spherical harmonics. Equation \ref{eqn:tidal-torque} is derived in appendix \ref{app:eom}, assuming a rigid asteroid and no third-body perturbations. Throughout the paper, we also refer to the ``inertial frame'' (the frame in which the orbit is fixed) and the ``body-fixed frame'' (the frame in which the asteroid is fixed), which are also discussed in appendix \ref{app:eom}.

In section \ref{sec:sim}, we describe the simulation used to integrate the equations of motion. In section \ref{sec:uncertainty}, we define the uncertainty model used to add noise to the data, and we describe how posterior probability distributions (PPDs) for density moments were extracted from spin-pole data in section \ref{sec:fit}. Finally, we outline a general model to extract density distributions from density moments in section \ref{sec:general-density} and define specific model implementations in section \ref{sec:specific-density}.

@@

We built a publicly accessible, custom simulation in C++ to produce angular velocity data as a function of time. This simulation requires as initial data (1) the orbital parameters of the asteroid $r_p$ (perigee distance) and $v_\infty$ (hyperbolic excess velocity); (2) the cadence of angular velocity observation $\Delta t$; (3) the central body moments $J_{\ell m}$, mass $\mu_\mathcal{B}$, and radius $a_\mathcal{B}$; (4) the initial asteroid angular velocity in the inertial frame $\bm \Omega_0$; (5) the asteroid length $a_\mathcal{A}$; and (6) the asteroid's density moments $K_{\ell m}$ and initial Euler angle $\gamma_0$. All parameters except (6) are assumed to be known to high accuracy. One can imagine that $a_\mathcal{A}$ is approximated by light-curve analysis, but if not, it is still necessary to fix $a_\mathcal{A}$ or else the values of $K_{\ell m}$ are degenerate with $a_\mathcal{A}$.

@@

With the simulation inputs specified, the equations of motion are integrated via the Runge-Kutta fourth order method, with a variable time step
\begin{equation}
  dt = dt_\text{min} + 10^{-3}(dt_\text{max} - dt_\text{min}) \brackets{\parens{\frac{D}{r_p}}^3 - 1}.
\end{equation}
The parameters $dt_\text{max}$ and $dt_\text{min}$ (20 and 10 seconds respectively) were chosen such that the numerical integration error was $\sim$100 times the floating point error, and that neighbouring values of $K_{\ell m}$ yielded significantly different spin pole data compared to floating point error. The data used to choose this $dt$ was obtained using the reference asteroid configuration, described in appendix \ref{app:reference-configs}.

@@

To add noise to data generated via the above simulation, we use the following uncertainty model. Each asteroid spin vector $\bm \Omega$ is assumed to be uncorrelated with other spin vectors, and we model uncertainty in the orientation and in the period as also uncorrelated. Consider a true spin vector $\bm \Omega^*$. For the sake of description, we work in coordinates in which $\bm \Omega^* \parallel \unit z$. Then, expressing the observed spin vector $\bm \Omega$ in spherical coordinates, we draw the polar angle from a normal distribution with standard deviation $\sigma_\theta$ centred on zero and the azimuthal angle from a uniform distribution. We also draw the ratio $\rho=\Omega/\Omega^*$ from a log-normal distribution centred on one, with width $\sigma_\rho$. This width is related to the more physical uncertainty in asteroid period $\sigma_P$ by $\sigma_P=P \sigma_\rho$ for small $\sigma_\rho$. Explicitly, the probability density function (PDF) of $\rho$ is 
\begin{equation}
  P(\rho) = \frac{1}{\rho\sqrt{2\pi \sigma_\rho^2}} \exp\parens{-\frac{\ln^2\rho}{2\sigma_\rho^2}}.
\end{equation}
See figure \ref{fig:uncertainty-model} for an illustration of the uncertainty model.

@@

  \draw[-{Latex[length=3mm]}] (0, 0) -- (-1.25, 3.35) node[anchor=south west] {$\bm \Omega$};
  \draw[rotate around={-50:(-3,3)}] (-3,3) ellipse (1.4 and 2);
  \draw (-3, 3) arc (110:93:6);
  \draw (-2, 3.5) node[anchor=center] {$\theta$};
  \draw (-0.5, 2.5) node[anchor=center] {$\rho \Omega^*$};

@@

\subsection{Extracting density moments from spin data}
\label{sec:fit}
Given synthetic data, an Affine Invariant MCMC Ensemble sampler was used to generate PPDs from flat priors. We use the Python implementation \texttt{emcee} \cite{foreman2013emcee}. Our parameters were $\gamma_0$, $K_{20}$, $K_{22}$, and $K_{3m}$ (10 in total), and were bounded by $|\gamma_0| < \pi/4$, and bounds on $K_{2 m}$ given in equation \ref{eqn:parameter-bounds}. Note that $\gamma_0$ is degenerate with $\gamma_0 + \pi/2$ since this is equivalent to re-labeling $\unit y$ as $\unit x$ and $\unit x$ as $-\unit y$. The other bounds were $|K_{3m}| < 1$. In general, spin data is most sensitive to $\gamma_0$ and $K_{2m}$, which we call the ``first-order parameters.'' We call $K_{3m}$ the ``second-order parameters.''

The MCMC was determined to converge when the fractional change in autocorrelation time (computed every 100 iterations) was one percent, and the number of iterations computed so far was more than 100 times the autocorrelation time. The MCMC fit also was set to terminate if more than $10^5$ iterations were run, but this only occurred for fits in which the data was too low-quality to resolve the parameters. $\sim 10^4$ iterations was often sufficient, which generally consumed about 7 hours of computation time on a super computer running 16 threads on 8 cores.

Before the MCMC was run, local minima in the likelihood were found via the Nelder-Mead algorithm implemented in \texttt{scipy} \cite{Gao2012}. Generally, only one local minimum existed, except when $K_{22}=0$ in which case rotational symmetry caused multiple values of $\gamma_0$ to be degenerate. Ensemble walkers were initialized near this local minimum, distributed by a Gaussian approximation of the likelihood, as determined via the Hessian of the likelihood at the minimum. Due to the high sensitivity of the angular velocity data to density moments, the minimization procedure sometimes failed to isolate the minimum likelihood. Therefore, a simpler simulation without the $K_{3m}$ terms of equation \ref{eqn:tidal-torque} was first used to minimize likelihood as a function of the first-order parameters $\gamma_0$ and $K_{2m}$, and then the full simulation was used to find the second-order parameters $K_{3m}$ with the first-order parameters fixed.

We further improved the minimization process by first minimizing with respect to data truncated at the point after perigee where $D=fr_p$ for some manually-set fraction $f$. In practice, $f=2$ often led to likelihood minima successfully found. The minimum was then further refined by minimizing based on the full data, with the previous minimum as the initial estimate.

@@

The asteroid density distribution $\rho_\mathcal{A}(\bm r)$ is not uniquely constrained via tidal torque interactions because only the density moments $K_{\ell m}$ contribute to equation \ref{eqn:tidal-torque}. However, by making sufficient assumptions about the density distribution, we can nevertheless estimate $\rho(\bm r)$ from $K_{\ell m}$. Here, we outline three different assumptions which constitute three models for extracting density distributions from density moments. All three of them are (partially) linear to facilitate computational ease and uncertainty propagation. One of them can be chosen, or another created, depending on what features are desired to be represented in the density distribution.

@@

We define a new coordinate system, the ``hybrid frame,'' which coincides exactly with body-fixed frame at the initial orientation of the asteroid assuming that the fit result for $\gamma_0$ is perfectly accurate. The orientation of the hybrid frame with respect to the inertial frame is therefore exactly known, so that $\mathds{1}$ is also exact in the hybrid frame and the body-fixed frame aligns with the hybrid frame up to uncertainty in $\gamma_0$. We will solve for $\rho(\bm r)$ in the hybrid frame given fit results $K_{\ell m}$ (known in the body-fixed frame) and the fixed $a_\mathcal{A}$ (equal in all frames).

The density moments defined in equation \ref{eqn:klm} are not linear in $\rho(\bm r)$, but we may fix $a_\mathcal{A}$ at the value assumed during the fit and set $\mu_\mathcal{A}$ equal to an arbitrary constant since the system is independent of the total asteroid mass. Thus, equations \ref{eqn:am} and \ref{eqn:klm} guarantees that $K_{\ell m}$ and $a_\mathcal{A}^2$ are linear in $\rho(\bm r)$. Equation \ref{eqn:am} and the $K_{00}=1$ component of equation \ref{eqn:klm} can each be applied as constraints to enforce these choices of $a_\mathcal{A}^2$ and $\mu_\mathcal{A}$.

Suppose that we restrict the number of degrees of freedom of $\rho(\bm r)$ from infinity to $m$ by explicitly defining some function $\rho(\bm r, \bm \Theta)$ for an $m$-dimensional vector $\bm \Theta$ which contains the free parameters of $\rho$. We will leave the explicit definition of $\rho(\bm r, \bm \Theta)$ to the model descriptions below, but for now we assume that $\rho$ is linear in $\bm \Theta$; i.e.,
\begin{equation}
  \rho(\bm r, \bm \Theta) = \bm B(\bm r) \cdot \bm \Theta
  \label{eqn:density-distro}
\end{equation}
for a $m$-dimensional vector $\bm B(\bm r)$. ($\bm B$ need not be linear in $\bm r$.) Thus, $a_\mathcal{A}^2$ and $K_{\ell m}$ are linear in $\bm \Theta$. We further assume that the model describes a way to reverse the relationship between $K_{\ell m}$ or $a_\mathcal{A}^2$ and $\bm \Theta$ so that
\begin{equation}
  \bm \Theta = A \bm K
  \label{eqn:density-model}
\end{equation}
where $A$ is a matrix. Here, we have arranged $a_\mathcal{A}^2$ and $K_{\ell m}$ into a vector $K$ which we say has $n$ dimensions. The order of this arrangement is irrelevant as long as it is kept consistent.

To propagate uncertainties from $\bm K$ to $\bm \Theta$ to $\rho(\bm r)$, we need the covariance matrix $\Sigma_K$ for $\bm K$. Suppose that the hybrid frame is offset from the body-fixed frame by some small angle $\Delta \gamma$, which results from uncertainty in $\gamma_0$. Then
\begin{equation}
  K_{\ell m}^\mathrm{hybrid} = e^{-im\Delta \gamma}K_{\ell m}^\mathrm{body-fixed}.
  \label{eqn:body-fixed-to-hybrid}
\end{equation}
by equation \ref{eqn:ylm-rotation}. Since $K_{\ell m}^\mathrm{body-fixed}$ was obtained by an MCMC fit, a large set of PPD-distributed samples is available for $\Delta \gamma$ and $K_{\ell m}^\mathrm{body-fixed}$, and the covariance matrix $\Sigma_K$ can be computed statistically in the hybrid frame by applying equation \ref{eqn:body-fixed-to-hybrid} to the samples. Then, propagation of uncertainty guarantees that the covariance matrix of $\bm \Theta$ is $\Sigma_\Theta = A \Sigma_K A^\dagger$ and the density distribution and uncertainty on density distribution are equal to
\begin{equation}
  \rho(\bm r) = \bm B(\bm r)^\dagger A\bm K \qquad \sigma^2_\rho(\bm r) = \bm B(\bm r)^\dagger A \Sigma_K A^\dagger \bm B(\bm r).
  \label{eqn:unc-rho}
\end{equation}
where $A^\dagger$ is the adjoint of $A$.

The role of a model is therefore to restrict the space of valid density distributions by defining the $m\times n$-dimensional constant matrix $A$ and the $m$-dimensional vector $\bm B(\bm r)$ such that equations \ref{eqn:density-model} and \ref{eqn:density-distro} are true. Then the density distribution and its uncertainty are given by equation \ref{eqn:unc-rho}. Three examples of possible models are given below.

\subsection{Model implementations for extracting density distributions from density moments}
\label{sec:specific-density}
